### 2024-04-09 paper/paper-papaja.Rmd
rm constraints on generality
revisionsand concept fluency stuff

abstract: |
  Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect. This basic feature of cognition may hamper efforts to curb misinformation: critical media consumption and fact-checking seem poised to fail if mere exposure can persuade people to adopt ideas. Under predominant accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we conducted three pre-registered experiments to examine whether mere exposure to statements produces genuine beliefs by examining whether people draw inferences from statements after mere exposure. Surprisingly, and in contrast to fluency or familiarity-based accounts of the illusory truth effect, we found that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements. This “illusory implication” effect suggests that the consequences of exposure to mis- and disinformation reach further than previously thought and calls for a new mechanistic account of these effects.

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "5209"

@@

The effect is one of "illusory" truth because it occurs following "mere exposure" to statements. That is, it occurs when the statements are seen or heard in a non-communicative context, such as when they are read during a true/false quiz. Generally, being told something---even by a source of unknown trustworthiness---is prima facie reason for believing it [@grice1989]. But reading a statement on a true/false quiz is not reason for believing it; the statement is just as likely to be false as to be true.

@@

Why should mere exposure affect later ratings of truth? Predominant theoretical accounts attribute illusory truth effects to metacognitive experiences of fluency or familiarity [@dechene.etal2010]. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [For a review, see @schwarz2015].

@@

@@

Both theoretical accounts provide a similar overall view of the illusory truth effect: Roughly, when asked whether something is true or false, people search their memory for knowledge pertaining to belief in the statement, but they also rely on their metacognitive experiences of familiarity and fluency. Familiarity and fluency are both signs that we have “heard this somewhere before.” According to foundational theories of communication, a general assumption that communicators strive to make true utterances is a prerequisite for successful communication and social functioning [@grice1989]. Thus, if having “heard something before” suggests that someone said it, then ecologically this is a reasonable cue to truth. Psychological studies that elicit the illusory truth effect hijack these metacognitive heuristics: They provide this sense of familiarity or fluency, but from a communicative context that lacks any reasonable assumption of truth.

@@

There is something unsettling about the illusory truth effect and the lack of agency it implies over our own beliefs. Even more worrisome, there could be important societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important consequences for the functioning of democratic society. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

@@

Generally, it is these wider consequences of belief (beyond truth ratings for a specific proposition) that should be most concerning. If illusory truth effects operate only through metacognitive influences, then these effects may not have much impact outside the lab.  Under these theoretical accounts, the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar statement. This would make the real-world consequences of the illusory truth effect far narrower than one might otherwise fear: These metacognitive mechanisms are liable to apply only in response to the exact statements previously seen, but are not as likely to generalize to other related ideas, or to drive actions or decision-making in novel contexts.

Nevertheless, concerns about how the influence of mere exposure might operate in contexts of pervasive misinformation [@difonzo.etal2016; @pennycook.etal2018] loom large enough to warrant further consideration. Could mere exposure to statements truly change beliefs?

@@

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether mere exposure affects beliefs is to examine whether mere exposure to statements can affect truth ratings for other related statements. We refer to this hypothetical impact of mere exposure on implied statements as the "Illusory Implication" Effect.

A handful of studies are suggestive in this regard: Unkelbach and Rom [-@unkelbach.rom2017] found that mere exposure to one statement could increase endorsements of implied statements. For instance, people were more likely to endorse the statement "Most accidents occur close to weekends" after exposure to the statement "Most accidents occur on Mondays". Similarly, Arkes and colleagues [-@arkes.etal1991] found that mere exposure to one set of statements also affected judgments about novel related statements on the same topic (China). However, in both cases these findings can also be easily explained by fluency or familiarity engendered by the similarity between the presented and tested statements [@arkes.etal1991; @unkelbach.rom2017].

A case where mere exposure lead to a lower truth rating for a related statement (i.e., one implied to be false) would more clearly implicate an effect of mere exposure on beliefs. In our review of the literature, we find three reported studies that offer preliminary evidence of this effect: Bacon [-@bacon1979] found that mere exposure to one statement ($P$) could lead to lower truth ratings for a new, contradictory statements($\neg P$) after a three-week delay. For instance, mere exposure to "crocodiles sleep with their eyes open" produced lower truth ratings for "crocodiles sleep with their eyes closed". Findings from an experiment by Garcia-Marques and colleagues [-@garcia-marques.etal2015] partially contradicted these results (using a slightly different exposure task), finding lower truth ratings for contradictory statements when participants were queried immediately, but finding the opposite pattern of results at the delay used by the original Bacon [-@bacon1979] study. A later series of experiments by Unkelbach and Rom [-@unkelbach.rom2017] generally replicated the newer findings [@garcia-marques.etal2015], although the effects were not observed consistently in each experiment.

@@

Here, we sought to expand upon prior experiments to decisively examine whether mere exposure to statements could influence people's beliefs by testing for an "illusory implication" effect. To do so, we examined whether people draw inferences from statements after exposure in a context without any presupposition of truth. We designed a series of studies to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for a  different statement the premise would logically entail (the “implication”). Crucially, we test this across cases where premises entail either the truth or falsehood of another related statement, ruling out purely metacognitive explanations.

In addition, we make several additional contributions toward definitively demonstrating the illusory implication effect beyond existing studies by 1) developing a new set of materials, 2) recruiting larger samples (present studies $N = 1020$, total prior cited studies $N = 452$), 3) utilizing modern hierarchical statistical modeling methods, and 4) taking great pains to ensure that participants' truth ratings are not influenced by task demands (Experiments 2 and 3). Across our studies, we find new compelling evidence for this “illusory implication” effect, indicating that mere exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

In Experiments 1 and 2, the effect of familiarity was weaker than the measured implication effect, so that average truth ratings were higher after exposure for false-implied-true items and lower for true-implied-false items (Figures 1 and 2). In Experiment 3 however, the estimated effect of familiarity and implication were similar, and truth ratings were similar for exposed and unexposed true-implied-false items (Figure 3). Although the estimate for the implication effect in our statistical model was clearly greater than zero, this may nevertheless allow a possible alternate explanation of our findings: that the true-implied-false items' premise statements simply do not engender as much familiarity or fluency for their implication as do the false-implied-true items, so that the effects differ between these groups. Put another way, the identification of the "implication" effect coefficient rests on the assumption that familiarity and fluency effects are not accidentally confounded with item groups.

@@

What cognitive mechanism could explain these findings? One potential explanation is that mere exposure to the premise statements leads participants to at least partially adopt them as beliefs. Simply entertaining the idea may create such an impression despite awareness of its ambiguous truth-value [@gilbert.etal1990; @gilbert1991; @garcia-marques.etal2015]. If so, this apparently occurs despite participant's awareness that the statements may be false, as ensured by constant reminders and comprehension checks in Experiment 3.

Alternately, the illusory implication effect may reflect explicit memory and source misattribution during testing: Participants may have some explicit memory for the premise statements, but fail to properly attribute their source to the experimental context. Future research might explore how explicit memory for the premise statements correlates with the strength of the illusory implication effects.

Whatever the cognitive mechanism behind the illusory implication effect we observed, these findings are further cause for concern about the spread and influence of misinformation. The spread of misinformation and fake news appears to outpace the spread of factual news [@vosoughi.etal2018] and creates massive challenges for online platforms seeking to remove, label, and correct misinformation [@sharma.etal2019]. These challenges may be further compounded by basic features of human cognition  that our studies have revealed: mere exposure to false statements can lead them to be _believed_.

## Constraints on Generality

The studies reported here targeted U.S. adults through an online survey provider (CloudResearch). Study materials were designed accordingly and may produce different results if tested on populations with different backgrounds and prior knowledge.

>>>>

abstract: |
  Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect.  Under predominant accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we consider whether exposure to “premise” statements affects people's truth ratings for novel “implied” statements, a pattern of findings a pattern of findings we call the "illusory implication" effect. We argue these effects would constitute evidence for genuine belief change from mere exposure and  identify a handful of existing findings that offer preliminary support for this claim. Building upon these, we conduct three new pre-registered experiments to further test this hypothesis, finding additional evidence that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements, including for considerably more distant implications than those previously explored. Our findings suggest that the effects of mere exposure reach further than previously thought, with potentially consequential implications for concerns around mis- and disinformation.

@@

keywords          : "Illusory truth; Metacognition; Cognitive Psychology; Misinformation"
wordcount         : "5636"

@@

The effect is one of "illusory" truth because it occurs following "mere exposure" to statements. That is, it occurs when the statements are seen or heard in a non-communicative context, such as when they are read during a true/false quiz. Generally, being told something---even by a source of unknown trustworthiness---is prima facie reason for believing it [@grice1989; also see @corneille.etal2020]. But reading a statement on a true/false quiz is not reason for believing it; the statement is just as likely to be false as to be true.

@@

Why should mere exposure affect later ratings of truth? Gilbert [-@gilbert1991] proposed that any idea entertained is by default first believed and must later be "unbelieved." However, predominant theoretical accounts attribute illusory truth effects not to belief but rather to metacognitive experiences of fluency or familiarity [@dechene.etal2010]. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [@alter.oppenheimer2009; @schwarz2015].

@@

Metacognitive fluency can derive from different sources, and several different types have been shown to influence ratings of truth. Fluency can be *perceptual*: statements printed more clearly or in easier-to-read fonts are rated as more true [@ReberSchwarz1999Effects]. Fluency can be influenced by *memory retrieval*, such that statements that are easier to retreive from memory are rated as more true [@BeggEtAl1992dissociation]. And fluency can also be semantic or *conceptual* in nature: concepts that are semantically primed are seen as better answers to trivia questions [@KelleyLindsay1993Remembering], and statements evoking previously-activated concepts are rated as more true [@silva.etal2017].

@@

Both theoretical accounts provide a similar overall view of the illusory truth effect: Roughly, when asked whether something is true or false, people search their memory for knowledge pertaining to belief in the statement, but they also rely on their metacognitive experiences of familiarity and fluency. Reliance on metacognitive experiences may be, at least to some extent, adaptive: Familiarity and fluency are both signs that we have “heard this somewhere before.” According to foundational theories of communication, a general assumption that communicators strive to make true utterances is a prerequisite for successful communication and social functioning [@grice1989]. Thus, if having “heard something before” suggests that someone said it, then ecologically this is a reasonable cue to truth. Psychological studies that elicit the illusory truth effect hijack these metacognitive heuristics: They provide this sense of familiarity or fluency, but from a communicative context that lacks any reasonable assumption of truth.

@@

There is something unsettling about the illusory truth effect and the lack of agency seems to imply about our own thinking. Even more worrisome, there could be important societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important consequences for the functioning of democratic society [also see @UnkelbachEtAl2019Truth]. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

@@

The extent to which the illusory truth effect operates entirely through metacognitive processes stands to determine the extent of its real-world consequences. Generally, it is the wider consequences of belief (beyond truth ratings for a specific proposition) that should be most concerning. If illusory truth effects operate only through metacognitive influences, then the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar situation or statement (which might include one activating similar concepts). To be sure, this might have important consequences, such as increasing the tendency to share information online after having been exposed previously [@vellani.etal2023]. Still, the impact of metacognitive effects seem less likely to drive inferences or decision-making in more novel situations. Instead, their influences would be constrained to operating within contexts similar to the initial exposure.

@@

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether mere exposure affects beliefs is to examine whether mere exposure to statements can affect truth ratings for other related or implied statements. We refer to this hypothetical impact of mere exposure on implied statements as the "Illusory Implication" Effect.

We believe a number of existing findings provide at least preliminary evidence for this illusory implication effect. A number of researchers have found that mere exposure affects truth ratings for paraphrases and direct implications or contradictions [@arkes.etal1991; @bacon1979; @garcia-marques.etal2015; @silva.etal2017; @unkelbach.rom2017]. A crucial question is how distantly inferences are drawn following mere exposure, and whether or not metacognitive accounts can explain these findings.

Silva and colleagues [-@silva.etal2017] found that mere exposure to statements increased truth ratings for paraphrases of those statements that expressed the same idea while sharing only minimal surface-level linguistic features. Conversely, exposure to statements *decreased* truth ratings for rephrased contradictions of those statements. Silva and colleagues [-@silva.etal2017] explain their findings as resulting from conceptual fluency and disfluency produced by the initial mere exposure to statements: prior exposure "activates" the relevant concepts, easing or interfering with their retrieval and use in processing subsequent statements and thereby producing different metacognitive experiences.

However, findings from Unkelbach and Rom [-@unkelbach.rom2017] cast doubt on a purely fluency-based account of these types of findings. These researchers also found that exposure to statements decreased truth judgments for contradictions of those statements. Crucially however, they also found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus "conceptual disfluency" is not likely to be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of mere exposure [@garcia-marques.etal2015; @gilbert1991; @unkelbach.rom2017].

@@

Prior work tested the effect of mere exposure on implied statements using only direct contradictions of the original statement (or paraphrases of direct contradictions). Here, we sought to further test for an "illusory implication effect" by testing the effect of mere exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).

We developed materials involving more distant inferences that do not afford direct one-to-one mappings between shared concepts in the premise and implication statements. For instance, one statement pair was the premise statement, “No U.S. astronauts have died since the Challenger explosion in 1986” paired with the implied statement, “The space shuttle Columbia disintegrated over Texas in 2003” (true, but implied to be false by the "premise"). In contrast with prior materials using "contradictory" statements, there is no direct one-to-one mapping of concepts between the two statements. Instead, a chain of inferences must developed for a conclusion to be drawn: that the space shuttle Columbia contained astronauts, that its disintegration would have killed them, and that 2003 is later than 1986.

In addition, we make several additional contributions toward definitively demonstrating the illusory implication effect beyond existing studies by 1) testing the effect with large samples of participants (present studies $N = 1020$), 2) utilizing modern hierarchical statistical modeling methods, and 3) taking great pains to ensure that participants' truth ratings are not influenced by task demands (Experiments 2 and 3). Across our studies, we find further compelling evidence for this “illusory implication” effect, indicating that mere exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

In Experiments 1 and 2, the effect of exposure was weaker than the measured implication effect, so that average truth ratings were higher after exposure for false-implied-true items and lower for true-implied-false items (Figures 1 and 2). In Experiment 3 however, the estimated effect of exposure and implication were similar, and truth ratings were similar for exposed and unexposed true-implied-false items (Figure 3). Although the estimate for the implication effect in our statistical model was clearly greater than zero, this may nevertheless allow a possible alternate explanation of our findings: that the true-implied-false items' premise statements simply do not engender as much familiarity or fluency for their implications as do the false-implied-true items, so that the effects differ between these groups. Put another way, the identification of the "implication" effect coefficient rests on the assumption that familiarity and fluency effects are not accidentally confounded with item groups.

@@

What cognitive mechanism could explain these findings? One potential explanation is that mere exposure to the premise statements leads participants to at least partially adopt them as beliefs. Simply entertaining the idea may create such an impression despite awareness of its ambiguous truth-value [@gilbert.etal1990; @gilbert1991; @garcia-marques.etal2015; @unkelbach.rom2017]. If so, this apparently occurs despite participant's awareness that the statements may be false, as ensured by constant reminders and comprehension checks in Experiment 3.

Modern theorists do not conceive of beliefs as lists of propositiosn, but instead as subserved by structured mental models or intuitive theories of how the world works [e.g. @gelman.legare2011; @gerstenberg.tenenbaum2017; @powell.etal2023; @johnson-laird2001; @thagard1989]. Thus, coming to believe a statement entails updating one's mental models with the new information conveyed by that statement. A key feature of these mental models is their coherence: updating them is expected to have wide-ranging and coherent implications [@gerstenberg.tenenbaum2017; @powell.etal2023; @thagard1989]. In the context of mere exposure effects, Unkelbach and Rom [-@unkelbach.rom2017] have proposed an account based on parallel constraint satisfaction within semantic networks, which might plausibly be elaborated to account for the present findings. In any case, we take the coherence of inferences following exposure to be evidence for a meaningful influence on people's mental models, and the key to our claim that mere exposure influences beliefs.

Finally, it is worth clarifying that our findings and wider claims about the illusory implication effect are broadly consistent with existing metacognitive explanations of the illusory truth effect: we hypothesize that the illusory implication effect is a distinct and somewhat orthogonal effect, based in appreciation of the meaning of statements rather than solely in the metacognitive experiences they produce. Nevertheless, it remains clear that fluency plays a crucial role in the classic illusory truth effect, especially as evidenced by neuroscientific findings [@wang.etal2016] and by the findings indicating the effect can be reversed when the pragmatics of the exposure context are changed [@corneille.etal2020].

## Conclusions

Whatever the cognitive mechanism behind the illusory implication effect we observed, these findings are further cause for concern about the spread and influence of misinformation. The spread of misinformation and fake news appears to outpace the spread of factual news [@vosoughi.etal2018] and creates massive challenges for online platforms seeking to remove, label, and correct misinformation [@sharma.etal2019]. These challenges may be further compounded by apparently basic features of human cognition: that mere exposure to false statements can lead them to be _believed_.
